---
title: "Demographic Influences on Municipal Budget Allocation and Infrastructure Development in Toronto"
subtitle: "Population density drive larger budget allocations meanwhile average household income drives construction projects"
author: 
  - Aviral Bhardwaj
thanks: "Code and data are available at: [https://github.com/Aviral-03/InfrastructureCausalModel-TO](https://github.com/Aviral-03/InfrastructureCausalModel-TO)." 
date: today
date-format: long
abstract: "This study examines the relationship between population density, average household income, and budget allocations across Toronto's 25 wards and investigates their combined influence on construction activity. Using data from the city's *Capital Budget Plan*, *Ward Profiles*, and *Active Building Permits*, we analyze spending patterns in key infrastructure-related areas. Employing a causal modeling approach, our findings reveal a negative association between population density and construction activity, despite some high-density wards receiving higher budget allocations. Additionally, average household income and total budget allocations exhibit a positive correlation with the number of building permits. These results underscore potential inequities in resource distribution, highlighting the need for data-driven and equitable urban budget planning to better address community needs."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
prefer-html: true
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(opendatatoronto)
library(readxl)
library(styler)
library(dplyr)
library(arrow)
library(ggrepel)
library(knitr)
library(kableExtra)
library(patchwork)
library(sf)
library(lintr)
library(DiagrammeR)
library(rsvg)
library(magrittr)
library(DiagrammeRsvg)
library(png)
library(webshot2)
library(gridExtra)
library(modelsummary)
library(rstanarm)
library(bayesplot)
```

```{r}
#| include: false
#| warning: false
#| message: false

# Get the analysis data
analysis_budget_data <- read.csv(here::here("data/02-analysis_data/analysis_budget_data.csv"))
analysis_building_permits <- read.csv(here::here("data/02-analysis_data/analysis_building_permits.csv"))
analysis_ward_data <- read.csv(here::here("data/02-analysis_data/analysis_ward_data.csv"))


# Get Model
model <- readRDS(here::here("models/model.rds"))
```

# Introduction {#sec-intro}

In urban governance, the equitable allocation of municipal budgets is essential to meet a city’s social, economic, and infrastructure needs. This challenge is especially pressing in Toronto, where 25 wards compete for limited resources amid a growing population [@Ontario2024DemographicTrends]. Understanding how demographic factors influence budget decisions remains under-explored. Therefore, this study seeks to answer: How do demographic factors such as population and income affect budget allocations across Toronto’s wards, and can we assess their combined impact on construction activity, a key indicator of infrastructure development?

Looking back, Toronto's historical budget data have shown uneven investment, particularly in high-density, lower-income areas, leading to disparities in infrastructure and services [@ferguson2024compromised]. However, following 2023 Mayor election, Mayor Olivia Chow promise to prioritize housing, transit, and equity-focused initiatives [@toronto_2024_budget]. The 2024 budget prioritizes significant investments in affordable housing, transportation, and community safety, while also focusing on long-term infrastructure development through a 10-year capital plan. Notable projects, such as the *Ontario Line* subway expansion, are projected to drive economic growth and urban development [@citeTransitExpansion]. Similarly, the *Sidewalks to Skylines* 10-year plan aims to create jobs and enhance residents' quality of life [@citeSidewalksToSkylines]. As these initiatives progress, ensuring equitable infrastructure investments will be crucial for Toronto’s sustainable growth. With the city’s population increasingly distributed across its wards, it is imperative to assess whether current budget allocations effectively align with broader development objectives amidst shifting demographic trends.

This study analyzes data from the City of Toronto’s Open Data Portal, incorporating demographic information from the *2021 Ward Profiles*, budget details from the *Capital Budget & Plan By Ward (10-Year Approved) 2021-2024*, and construction data from *Building Applications and Permits*. Using a Bayesian multi-regression model guided by a causal framework, we found that both average household income and total budget allocations significantly influence the number of building permits. However, population density displayed an unexpected negative relationship with construction activity, suggesting the presence of unobserved factors affecting this link. To address this, we excluded population density as an explanatory variable and confirmed that higher-income wards receiving larger budget allocations corresponded to more construction activity. These findings highlight the importance of understanding how socioeconomic and budgetary factors shape urban development, emphasizing the need for equitable resource distribution to support balanced growth across Toronto’s wards.

The paper is organized as follows: [@sec-data] outlines the data sources and methodology used in the analysis. [@sec-model] explains the Bayesian model applied to estimate the effects of demographic and economic factors, following the causal model ([@fig-causal-model]). [@sec-results] presents key findings, including trends in budget allocations, the relationship between demographic factors and budget distribution, and the impact on construction projects. [@sec-discussion] explores the implications for urban governance and budget planning in Toronto. Finally, [@sec-limitations] addresses the study’s limitations and offers recommendations for future research. Additional details, including tables, figures, and methodology, are provided in the [@sec-appendix].

## Estimand

Our research investigates the causal relationships between population, income, budget allocation, and active construction projects, as illustrated in [@fig-causal-model]. Population and income are key confounders, influencing both budget allocation and the number of construction projects. Budget allocation serves a dual role: as a mediator connecting population and income to construction projects and as an instrumental variable, helping to estimate their effects on construction activity. The outcome variable is the number of active construction projects, while population and income are the primary predictors. Our goal is to estimate the causal effect of budget allocation on construction activity, accounting for the confounding effects of population and income.

Population is a important demographic indicator, representing the number of residents in each ward, while average household income reflects economic well-being, influencing access to resources and quality of life [@Schaeffer]. Therefore, these two variables of interest were chosen for the analysis. For budget allocations we choose these specific categories of interest, specifically Growth-Related expenditures, State of Good Repair, and Service Improvement and Enhancement.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-causal-model
#| fig-cap: "Causal model: Population and income influence budget allocation and construction projects"
#| fig-pos: "h"
#| fig.align: "center"

# create a dag for the model
causal_model <- "digraph {
  graph []
  # Nodes
  node [shape = plaintext]
    A [label = 'Population']
    B [label = 'Budget Allocation']
    C [label = 'Active Construction Projects']
    D [label = 'Income']
{ rank = same A D; }
  # Edges for confounding variable
  edge []
    A -> {B C}
    D -> {B C}
    B -> C
}"

# render the causal model
exporting_image <- grViz(causal_model) |>
  export_svg() |>
  charToRaw() |>
  rsvg_png(here::here("other/sketches/causal_model.png"), width = 800, height = 600)

image <- readPNG(here::here("other/sketches/causal_model.png"))
par(mar = rep(2, 4))
plot.new()
rasterImage(image, 0, 0, 1, 1)  
```


# Data {#sec-data}

The raw data was sourced from the City of Toronto's Open Data Portal using the `opendatatoronto` [@openDataToronto] package. For the purpose of this analysis we used several data sets: `2023 Ward Profiles (25-Ward Model)` [@toronto_ward_profiles], `Capital Budget and Plan Details from 2021-2024` [@toronto_budget_plan], `City Wards` [@toronto_city_wards], and `Building Permits - Active Permits` [@toronto_building_permits].

The data, provided in CSV formats, was cleaned and analyzed using R [@citeR] programming language. The `readxl` [@readxl] package was used for reading Excel files. Other R packages used includes `tidyverse` [@tidyverse], `styler` [@styler], and `dplyr` [@dplyr] for creating tables. `ggplot2` [@ggplot2] and `kableExtra` [@kableExtra] were used for data visualization and table formatting. The `patchwork` [@patchwork] package was used for combining multiple plots, and `sf` [@sf] for spatial data analysis. For model estimation, the `rstanarm` [@citerstanarm] package was used for Bayesian modeling, and `bayesplot` [@citeBayesPlot] for visualizing the results. The `lintr` [@citeLintr] package was used for code linting.

## Ward Profiles (25-Ward Model) {#sec-ward-profiles}

The 2021 Ward Profiles [@toronto_ward_profiles], based on the 25-Ward model were provided by City Planning. These profiles included census data from the 2021, 2016, and 2011 Census of Population, covering demographic, social, and economic information for each ward in Toronto. These variables were collected through methods including online responses, mailed questionnaires, the Census Help Line, and enumerators [@nhs_canada].

These questionnaires gathered information on various topics related to residents' demographic characteristics, such as education, household income, number of dependents, employment status and etc. Participation in the survey is voluntary, and data is collected directly from residents, including their postal codes, which are used to determine their respective wards. To ensure privacy and confidentiality, the data is subsequently aggregated and anonymized [@canada_statcan_2023_nhs].

This data-set was included in this analysis to provide understanding into the population and average household income for each ward, providing understanding into the city's socioeconomic landscape. 25-Ward model was used instead of the 44-Ward model as it was the most recent data available at the time of analysis and matched the Capital Budget data.

The data was stored in an Excel workbook with multiple tabs, but for this analysis, we used the first tab, `2021 Census One variable`, which contains data for all 25 wards (Ward 1, Ward 2, ..., Ward 25). After cleaning, the data was saved in CSV and Parquet formats, with the following columns:

-   `ward_id`: unique identifier for each ward,
-   `ward`: ward name,
-   `population`: total population,
-   `income`: average household income.

The ward names were manually entered into the cleaned data to match with `ward_id`. A sample of the data can be seen in [@tbl-ward-profile-data].

```{r}
#| echo: false
#| message: false
#| label: tbl-ward-profile-data
#| tbl-cap: Sample of Cleaned Toronto Ward Profile Data
#| tbl-pos: "h"

# Display the first few rows of the cleaned ward profile data using kable
analysis_ward_data |>
  head() |>
  kable(col.names = c("Ward ID", "Ward Name", "Population", "Income")) |>
  kable_styling(full_width = F, position = "center")
  
```

## Capital Budget and Plan Details {#sec-capital-budget}

Each year, the City of Toronto publishes the Capital Budget and Plan Details dataset [@toronto_budget_plan], which outlines a 10-year capital budget and plan. This dataset breaks down the capital budget across the city’s 25 wards, allocating funds for infrastructure projects, equipment purchases, and other fixed assets. This budget is developed through a collaborative process, where city staff prepare an initial draft, which is then reviewed by the Budget Committee. Input is solicited from Toronto residents and businesses, and subsequently, the Mayor presents the finalized budget proposal by February 1. City Council reviews and considers this budget within 30 days [@toronto_2024_budget].

For the purpose of this analysis we selected the year 2021-2024 to align with the 2022 municipal elections and the subsequent relevance to planning efforts. Furthermore, the city's budgeting process underwent significant shifts after 2020 due to the impact of the COVID-19 pandemic, which altered spending priorities and resource allocation. Focusing on the 2021–2024 timeframe allows us to analyze the post-pandemic period, avoiding the uncertainties of the pandemic and ensuring the data remains consistent and reliable.

Each budget plan includes five primary categories under *State of Good Repair*, *Growth Related*, *Health and Safety*, *Service Improvement and Enhancement*, and *Legislated*. These categories define the main areas where capital expenditures are directed. For this analysis, however, we will focus on these three variables of interest: *State of Good Repair*, *Growth Related*, and *Service Improvement and Enhancement*.

Since our analysis aims to understand how budget allocations influence construction activity and infrastructure development, we selected these three categories as they are most relevant to capital expenditures and infrastructure projects. *State of Good Repair* focuses on maintaining and preserving existing infrastructure, *Growth Related* addresses the expansion and development of new infrastructure, and *Service Improvement and Enhancement* aims to enhance public services and amenities.

Raw data includes key columns such as `Project Name`, yearly budget allocations for each year, `Ward Number`, `Ward`, `Category`, and `Total 10 Year` (Sum of Year 1 to 10), where the budget is in thousands of dollars (e.g., 10 = \$10,000). [@tbl-capital-budget-data] shows a sample of the cleaned data along with our variables of interest:

-   `Ward ID`
-   `Ward Name`
-   `Category of the Capital Budget`
-   `Total 10-year capital budget` allocated to each ward

Rows with `CW` (city-wide budget) were removed since they were applicable to all wards.

```{r}
#| echo: false
#| message: false
#| label: tbl-capital-budget-data
#| tbl-cap: Sample of Cleaned Toronto Capital Budget Data
#| tbl-pos: "h"

analysis_budget_data |>
  head() |>
  select(ward_id, ward, category, total_10_year) |>
  kable(col.names = c("Ward ID", "Ward Name", "Category", 
                  "Total 10-Year Budget (in 000s)"))
```

## City Wards {#sec-ward-names}

The City Wards dataset [@toronto_city_wards], published by the City Clerk's Office and last updated on July 22, 2024, contains geographical information about each ward, including the ward ID, ward name, and ward boundary. These ward boundaries were decided as a part of *Bill 5, Better Local Government Act* in 2018, reducing the number of wards from 47 to 25 [@toronto_city_wards].

This dataset, effective January 1, 2024, was used to map the `ward_id` to the ward name in the cleaned data. Key columns include:

-   `ward_id`: unique identifier for each ward,
-   `ward`: ward name,
-   `ward_boundary`: geographical boundary of the ward.

The ward names were mapped to the `ward_id` and integrated with the Ward Profiles [@sec-ward-profiles] and Capital Budget data sets [@sec-capital-budget] to create the final data set for analysis. This dataset was not used directly in the analysis but was essential for mapping the ward names to the ward IDs in the cleaned data.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-toronto-map-population
#| fig-cap: "Map of Toronto highlighting the population and income densities by ward"
#| fig.pos: 'H'
#| fig.height: 5

# URL to the zip file
url <- "https://ckan0.cf.opendata.inter.prod-toronto.ca/dataset/5e7a8234-f805-43ac-820f-03d7c360b588/resource/35f67d86-cfc8-4483-8d77-50d035b010d9/download/25-ward-model-december-2018-wgs84-latitude-longitude.zip"

# Temporary file to store the downloaded zip
temp_zip <- tempfile(fileext = ".zip")

# Download the zip file
download.file(url, temp_zip, mode = "wb")

# Unzip the file to a temporary directory
temp_dir <- tempdir()
unzip(temp_zip, exdir = temp_dir)

shapefiles <- list.files(temp_dir, pattern = "\\.shp$", full.names = TRUE)

toronto_map <- st_read(shapefiles[1], quiet = TRUE)

# Clean up the downloaded zip file
unlink(temp_zip)

# Merge analysis_data with toronto_map by ward_id
toronto_map <- merge(toronto_map, analysis_ward_data, 
                     by.x = "AREA_NAME", by.y = "ward") |>
  rename("income" = "average_household_income")

color_scale <- scale_fill_gradient(low = "lightblue", high = "darkblue")


# Plot with facet_wrap for average household income and population
plot1 <-ggplot() +
  geom_sf(data = toronto_map, aes(fill = population)) +
  geom_sf_text(data = toronto_map, aes(label = ward_id), 
               size = 2, color = "white") +
  color_scale +
  theme_minimal() +
  labs(x = "Longitude", y = "Latitude") +
  theme(legend.position = "right")


plot2 <- ggplot() +
  geom_sf(data = toronto_map, aes(fill = income)) +
  geom_sf_text(data = toronto_map, aes(label = ward_id), size = 2, color = "white") +
  color_scale +
  theme_minimal() +
  labs(x = "Longitude", y = "Latitude") +
  theme(legend.position = "right")

plot1 / plot2


```

## Building Permits - Active Permits {#sec-building-permits}

The *Building Permits - Active Permits* dataset [@toronto_building_permits], published by the City Planning Division and last updated on November 28, 2024, serves as an important record of active building permits in Toronto. A building permit is a municipally issued document, mandated by the Building Code Act and enforced by the City of Toronto, regulating the construction or demolition of physical structures [@toronto_building_permits].

The process of obtaining a building permit involves submitting an application to the City of Toronto, including necessary drawings, documents, and other forms based on the permit type. The Building Division reviews the application, and a Toronto Building Inspector ensures compliance with the Ontario Building Code, Zoning By-law, and other applicable regulations. Once approved, the permit is issued, allowing the applicant to commence construction or demolition, and thus data is saved in the system as an active permit.

[@tbl-raw-building-permits-data] outlines detailed information about active building permits in Toronto, including key features such as `Permit Number`, `Permit Type`, `Structure Type`, `Status`, and more.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-raw-building-permits-data
#| tbl-cap: "Sample of the raw building permits data"
#| tbl-pos: "h"

# Load the raw building permits data
building_permits <- read_csv(here::here("data/01-raw_data/building_permits_data.csv"))

# Display the first few rows of the building permits data using kable
building_permits |>
  select("_id", PERMIT_TYPE, STRUCTURE_TYPE, WORK, POSTAL, STATUS) |>
  head() |>
  kable(col.names = c("ID", "Permit Type", "Structure Type", "Work", "Postal Code", "Status"))
```

We selected this data-set to evaluate how our confounders, population and income, influence construction activity, and mediator, budget allocation, correlate with the number of building permits issued in each ward. The number of permits serves as a proxy for construction activity and development, highlighting the demand for infrastructure investment and capital expenditures. By examining the relationship between building permits and outlined variables, we can understand how construction activity influences resource distribution across the city's wards.

From the raw data, we selected key variables of interest: `Postal Code`, `Status`, and `Work`. The `Postal Code` was used to map building permits to their respective wards, while `Status` and `Work` were utilized to filter permits based on their current state and type of work. [@tbl-status-building-permits-data] lists 45 unique statuses assigned to each submitted permit. For this analysis, we focused on permits with the following statuses: `Approved`, `Application Accepted`, `Issuance Pending`, `Ready for Issuance`, `Permit Issued`, and `Permit Issued/Close File`. These statuses indicate that the permit has been approved and that construction or demolition is either underway or completed.

<!-- To ensure we analyzed only construction-related permits, we filtered out entries where `Work` did not have the value `New Building`. -->
The cleaned data was then saved in CSV and Parquet formats with the following columns:

-   **`ward_id`**: Unique identifier for each ward.\
-   **`total_building_permits`**: Total number of building permits issued in each ward.

[@fig-building-permits-data] illustrates the total number of building permits by ward, providing understanding into construction activity and development patterns across Toronto’s wards.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-building-permits-data
#| fig-cap: "Total number of building permits by Ward"
#| fig-pos: "h"
#| fig-height: 5
#| fig-width: 10

# Display the total number of building permits by ward using kable
analysis_building_permits |>
  ggplot(aes(x = ward_id, y = total_building_permits)) +
  geom_histogram(stat = "identity", fill = "skyblue", color = "darkblue") +
  labs(x = "Ward ID", y = "Number of building permits") +
  scale_x_continuous(breaks = seq(1, 25, 1)) +
  # change the theme
  theme_minimal() +
  scale_fill_brewer(palette = "Set2")
```

## Trends in Budget Allocations (2021-2024) {#sec-budget-trends}

```{r}
#| echo: false
#| warning: false
#| message: false

# select only growth related
growth_related <- analysis_budget_data |>
  filter(category == "Growth Related") |>
  mutate(total_10_year = total_10_year / 1000) |>
  # remove negative values
  filter(total_10_year > 0) |>
 rename(total_budget = total_10_year)

# Calculate the outliers using IQR
outliers <- growth_related |>
  group_by(year) |>
  mutate(
    Q1 = quantile(total_budget, 0.25),
    Q3 = quantile(total_budget, 0.75),
    IQR = Q3 - Q1,
    is_outlier = total_budget < (Q1 - 1.5 * IQR) | total_budget > (Q3 + 1.5 * IQR)
  ) |>
  filter(is_outlier)

# Plot with labels for outliers
plot1 <- ggplot(growth_related, aes(x = factor(year), y = total_budget)) +
  geom_boxplot(fill = "skyblue", color = "darkblue") +
  geom_jitter(aes(color = as.factor(ward_id)), width = 0.2) +
  geom_text(
    data = outliers, 
    aes(label = ward), 
    nudge_y = 15, 
    check_overlap = TRUE
  ) +  # Label only outliers
  labs(
    x = "Year",
    y = "Total Budget (in thousands)"
  ) +
  theme_minimal() +
  # remove legend
  theme(legend.position = "none")
```

```{r}
#| echo: false
#| warning: false
#| message: false

# select only growth related, state of good repair and service improvement and enhancement
state_of_good_repair <- analysis_budget_data |>
  filter(category == "State of Good Repair") |>
  mutate(total_10_year = total_10_year / 1000) |>
  # remove negative values
  filter(total_10_year > 0) |>
 rename(total_budget = total_10_year)

# Calculate the outliers using IQR
outliers <- state_of_good_repair |>
  group_by(year) |>
  mutate(
    Q1 = quantile(total_budget, 0.25),
    Q3 = quantile(total_budget, 0.75),
    IQR = Q3 - Q1,
    is_outlier = total_budget < (Q1 - 1.5 * IQR) | total_budget > (Q3 + 1.5 * IQR)
  ) |>
  filter(is_outlier)

# Plot with labels for outliers
plot2 <- ggplot(state_of_good_repair, aes(x = factor(year), y = total_budget)) +
  geom_boxplot(fill = "skyblue", color = "darkblue") +
  geom_jitter(aes(color = as.factor(ward_id)), width = 0.2) +
  geom_text(
    data = outliers, 
    aes(label = ward), 
    nudge_y = 15, 
    check_overlap = TRUE
  ) +  # Label only outliers
  labs(
    x = "Year",
    y = "Total Budget (in thousands)"
  ) +
  theme_minimal() +
  # remove legend
  theme(legend.position = "none")
```

```{r}
#| echo: false
#| warning: false
#| message: false
# select only growth related, state of good repair and service improvement and enhancement
service_improvement <- analysis_budget_data |>
  filter(category == "Service Improvement and Enhancement") |>
  mutate(total_10_year = total_10_year / 1000) |>
  # remove negative values
  filter(total_10_year > 0) |>
 rename(total_budget = total_10_year)

# Calculate the outliers using IQR
outliers <- service_improvement |>
  group_by(year) |>
  mutate(
    Q1 = quantile(total_budget, 0.25),
    Q3 = quantile(total_budget, 0.75),
    IQR = Q3 - Q1,
    is_outlier = total_budget < (Q1 - 1.5 * IQR) | total_budget > (Q3 + 1.5 * IQR)
  ) |>
  filter(is_outlier)

# Plot with labels for outliers
plot3 <- ggplot(service_improvement, aes(x = factor(year), y = total_budget)) +
  geom_boxplot(fill = "skyblue", color = "darkblue") +
  geom_jitter(aes(color = as.factor(ward_id)), width = 0.2) +
  geom_text(
    data = outliers, 
    aes(label = ward), 
    nudge_y = 15, 
    check_overlap = TRUE
  ) +  # Label only outliers
  labs(
    x = "Year",
    y = "Total Budget (in thousands)"
  ) +
  theme_minimal() +
  # remove legend
  theme(legend.position = "none")
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-budget-trends
#| fig.cap: "Trends in budget allocations (2021-2024) by category"
#| fig-subcap: ["Growth Related", "State of Good Repair", "Service Improvement and Enhancement"]
#| fig-pos: "h"
#| layout-ncol: 2

plot1
plot2
plot3
```

The box plots in [@fig-budget-trends] illustrate the trends in budget allocations across three categories—*Growth Related*, *State of Good Repair*, and *Service Improvement and Enhancement*—from 2021 to 2024. Each box plot represents the distribution of budget allocations for each year, with the median value indicated by the horizontal line inside the box. Outliers are labeled with the corresponding ward ID, highlighting wards with unusually high or low budget allocations.

Ward 10 (Spadina-Fort York) received the highest budget allocations across both *Growth Related* and *State of Good Repair* categories, indicating significant investment in infrastructure development and maintenance to population-dense areas. Toronto Center (Ward 13), Don Valley East (Ward 16), and Etobicoke Center (Ward 2) also received notable budget allocations, some of these wards have higher average household incomes as well as higher population densities. The box plots reveal that budget allocations for *Growth Related* projects have increased steadily over the years, with 75% of wards receiving higher allocations in 2024 compared to 2021. In contrast, budget allocations for *State of Good Repair* and *Service Improvement and Enhancement* projects have remained relatively consistent, with few outliers receiving higher budget allocations.

The box plots for *State of Good Repair* and *Service Improvement and Enhancement* projects however, are less varied as most of the wards received similar budget allocations across the years, with few outliers such as University-Rosedale, Toronto Centre and Don Valley West receiving higher budget allocations, who coincidentally have higher average household income.

## Relationship between Variables of Interest and Budget Allocations {#sec-rel-variables}

Our budget data spans three categories—*Growth Related*, *State of Good Repair*, and *Service Improvement and Enhancement*—for the years 2021-2024. As observed in [@sec-budget-trends], there has been a steady increase in budget allocations across these categories. To better understand the relationship between these budget categories and our variables of interest, we calculate the average budget allocation for each category across the years. The results are presented in [@tbl-avg-budget-data].

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-avg-budget-data
#| tbl-cap: "Average 10-year budget allocations by category (2021-24)"
#| tbl-pos: "h"

# Calculate the average 10-year budget allocations by category
average_budget <- analysis_budget_data |>
  group_by(ward_id, ward, category) |>
  summarise(avg_total_10_year = mean(total_10_year)) |>
  pivot_wider(names_from = category, values_from = avg_total_10_year) |>
  select(ward_id, ward, `Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`)

average_budget |>
  head() |>
  kable(col.names = c("Ward ID", "Ward Name", "Growth Related", 
                      "State of Good Repair", "Service Improvement and Enhancement"))
```

### Relationship between Confounders and Budget Allocations (Service Improvement and Enhancement)
```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-service-improvement-budget
#| fig.cap: "Service Improvement and Enhancement Budget by Ward"
#| fig-pos: "h"
#| fig.width: 10
#| fig.height: 5


combined_data <- analysis_ward_data |>
  left_join(average_budget, by = "ward_id") |>
  left_join(analysis_building_permits, by = "ward_id") |>
  rename(ward = ward.x) |>
  select(ward_id, ward, population, average_household_income, `Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`, total_building_permits)


s_i_budget <- combined_data |>
  select(ward_id, ward, population, average_household_income, `Service Improvement and Enhancement`)

p1 <- ggplot(s_i_budget, aes(x = population, y = `Service Improvement and Enhancement`, label= ward_id)) +
  geom_point() +
  geom_text_repel(hjust = -0.5, vjust = 0.5) +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  labs(x = "Population", y = "Average Total 10-Year Budget (in 000s) 2021-24") +
  theme_minimal()

p2 <- ggplot(s_i_budget, aes(x = average_household_income, y = `Service Improvement and Enhancement`, label= ward_id)) +
  geom_point() +
  geom_text_repel(hjust = -0.5, vjust = 0.5) +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  labs(x = "Average Household Income", y = "") +
  theme_minimal()

# Combine both plots side by side
grid.arrange(p1, p2, ncol = 2)
```
[@fig-service-improvement-budget] illustrates the relationship between average household income, population, and the total budget allocation each year (2021-2024) for *Service Improvement and Enhancement* projects. These projects  aim to enhance the quality of services provided to residents and improve the overall infrastructure of the city [@toronto_improvements].

The linear regression lines for average household income and population shows a positive correlation with budget allocation for Service Improvement and Enhancement projects. Wards with higher average household incomes, such as Ward 11 (University-Rosedale) receive a larger share of the budget allocation for these projects, while wards with higher population densities, such as Ward 10 (Spadina-Fort York) and Ward 3 (Etobicoke-Lakeshore), also receive a proportionally larger share of the funds. This suggests that both income levels and population density play a significant role in determining funding levels for Service Improvement and Enhancement projects, with higher-income and more populous wards benefiting from increased allocations.

Other wards like Ward 15 which has the highest average household income, but is one of least populous wards, receives a lower budget allocation for Service Improvement and Enhancement projects, indicating that population density may be a more significant factor in determining funding levels than income levels.

Despite these outliers, the majority of wards received similar budget allocations for Service Improvement and Enhancement projects, underscoring a city-wide strategy to enhance services uniformly for all residents.

### Relationship between Confounders and Budget Allocations (Growth-Related and State of Good Repair)

[@fig-growth-related-budget]a) illustrates the relationship between average household income, population, and the total budget allocation for *Growth-Related projects*. The linear regression line for average household income remains relatively flat, hovering around \$100,000. This suggests that budget allocations for Growth-Related projects are not significantly influenced by income levels, indicating a more uniform distribution of funds across wards regardless of household income. A similar lack of correlation is observed in [@fig-growth-related-budget]b), where budget allocations for State of Good Repair projects also show minimal dependence on average household income.

In contrast, the linear regression line for population shows a strong positive correlation with budget allocation for Growth-Related projects. Wards with higher population densities—such as Ward 10 (Spadina-Fort York) and Ward 3 (Toronto Centre)—receive a larger share of the Growth-Related budget. This indicates that population density plays an important role in determining funding levels for these projects, with more populous wards benefiting from increased allocations. A similar trend is observed in the budget distribution for State of Good Repair projects, where wards with higher population densities receive a proportionally larger share of the funds.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-growth-related-budget
#| fig.cap: "Growth and State of Good Repair Budget by Ward"
#| fig-subcap: ["Growth Related", "State of Good Repair"]
#| fig-pos: "h"
#| fig.width: 10
#| fig.height: 3

growth_budget <- combined_data |>
  select(ward_id, ward, population, average_household_income, `Growth Related`)

p1 <- ggplot(growth_budget, aes(x = population, y = `Growth Related`, label= ward_id)) +
  geom_point() +
  geom_text_repel(hjust = -0.5, vjust = 0.5) +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  labs(x = "Population", y = "Average Total 10-Year Budget (in 000s) 2021-24") +
  theme_minimal()

p2 <- ggplot(growth_budget, aes(x = average_household_income, y = `Growth Related`, label= ward_id)) +
  geom_point() +
  geom_text_repel(hjust = -0.5, vjust = 0.5) +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  labs(x = "Average Household Income", y = "") +
  theme_minimal()
good_repair_budget <- combined_data |>
  select(ward_id, ward, population, average_household_income, `State of Good Repair`)



p3 <- ggplot(good_repair_budget, aes(x = population, y = `State of Good Repair`, label= ward_id)) +
  geom_point() +
  geom_text_repel(hjust = -0.5, vjust = 0.5) +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  labs(x = "Population", y = "Average Total 10-Year Budget (in 000s) 2021-24") +
  theme_minimal()

p4 <- ggplot(good_repair_budget, aes(x = average_household_income, y = `State of Good Repair`, label= ward_id)) +
  geom_point() +
  geom_text_repel(hjust = -0.5, vjust = 0.5) +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  labs(x = "Average Household Income", y = "") +
  theme_minimal()

# Combine both plots side by side
grid.arrange(p1, p2, ncol = 2)
grid.arrange(p3, p4, ncol = 2)

# | echo: false
# | message: false
# | warning: false
# | label: fig-good-repair-budget
# | fig.cap: "State of Good Repair Budget by Ward"
# | fig-pos: "h"
# | fig.width: 10
# | fig.height: 5
```


# Model {#sec-model}

The purpose of this paper is to investigate the relationship between population density, average household income, and budget allocations across Toronto's 25 wards, and their potential impact on the economic well-being of each ward, as represented by the `total_building_permits` issued, a indicator of infrastructure development. To achieve this, we will fit a Bayesian regression model with varying intercepts to estimate the effect of these variables on the number of building permits issued in each ward.

## Model set-up

The particular model we used is Bayesian multiple linear regression model with varying intercepts. The model includes the following variables:

\begin{align*} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \beta_0 + \beta_1 x_{\text{income}} + \beta_2 x_{\text{budget}} + \gamma_i \\
\beta_0 &\sim \mbox{Normal}(0, 2.5) \\
\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
\beta_2 &\sim \mbox{Normal}(0, 2.5) \\
\gamma_i &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align*}

In the above model:

- $\mu_i$ is the expected number of building permits issued in ward $i$.
- $\beta_0$ is the intercept, representing the expected number of building permits issued in a ward with average values for all other variables.
- $\beta_1$ is the coefficient for the predicted change in the number of building permits issued in a ward given a one unit increase in the average household income of the ward.
- $\beta_2$ is the coefficient for the predicted change in the number of building permits issued in a ward given a one unit increase in the total budget allocation across three categories: Growth Related, State of Good Repair, and Service Improvement and Enhancement.

The model also includes a varying intercept $\gamma_i$ for each ward, accounting for the unobserved heterogeneity between wards. The varying intercepts allow the model to capture the unique characteristics of each ward that may influence the number of building permits issued.

The model was fitted using the `stan_glmer` function from the `rstanarm` package [@citerstanarm] in R [@citeR] and `modelsummary` [@citeModelsummary] for model summary tables. The prior distributions for the coefficients were set to normal distributions with a mean of 0 and a standard deviation of 2.5. The model also included a prior for the intercept, which was set to a normal distribution with a mean of 0 and a standard deviation of 2.5. The standard deviation of the likelihood was set to an exponential distribution with a rate parameter of 1.

\newpage

## Model justification

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-model-justification
#| tbl-cap: "Mean, Median, and Standard Deviation of Each Variable"
#| tbl-pos: "h"

results_data <-
  analysis_budget_data |>
  group_by(ward_id, ward, category) |>
  summarise(total_10_year = mean(total_10_year)) |>
  pivot_wider(names_from = category, values_from = total_10_year) |>
  select(ward_id, ward, `Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`) |>
  left_join(analysis_building_permits, by = "ward_id") |>
  left_join(analysis_ward_data, by = "ward_id") |>
  rename(ward = ward.x) |>
  select(ward_id, ward, population, average_household_income, total_building_permits, `Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`) |>
  # add a new column for total budget across three categories
  mutate(total_budget = `Growth Related` + `State of Good Repair` + `Service Improvement and Enhancement`) |>
  select(-`Growth Related`, -`State of Good Repair`, -`Service Improvement and Enhancement`)

# mean of all population, average household income, total budget and total building permits
mean_population <- mean(results_data$population)
mean_income <- mean(results_data$average_household_income)
mean_budget <- mean(results_data$total_budget)
mean_building_permits <- mean(results_data$total_building_permits)

# median of all population, average household income, total budget and total building permits
median_population <- median(results_data$population)
median_income <- median(results_data$average_household_income)
median_budget <- median(results_data$total_budget)
median_building_permits <- median(results_data$total_building_permits)

# standard deviation of all population, average household income, total budget and total building permits
sd_population <- sd(results_data$population)
sd_income <- sd(results_data$average_household_income)
sd_budget <- sd(results_data$total_budget)
sd_building_permits <- sd(results_data$total_building_permits)

# Create a data frame with the mean, median, and standard deviation of each variable
model_justification <- data.frame(
  Variable = c("Population", "Average Household Income", "Total Budget", "Total Building Permits"),
  Mean = c(mean_population, mean_income, mean_budget, mean_building_permits),
  Median = c(median_population, median_income, median_budget, median_building_permits),
  SD = c(sd_population, sd_income, sd_budget, sd_building_permits)
)

# Kable table
model_justification |>
  kable(col.names = c("Variable", "Mean", "Median", "Standard Deviation"))

```

To ensure model stability and interpretability, we standardized predictor variables (population density, average household income, and budget allocations) by subtracting their means and dividing by their standard deviations, as detailed in [@tbl-model-justification]. This step addressed differences in scale and prevented numerical instability during Bayesian multiple linear regression modeling. 

The Bayesian multiple linear regression model was selected because the outcome variable—building permits—is continuous rather than count-based, making Poisson or negative binomial models unsuitable. The chosen predictors reflect demographic, economic, and infrastructural factors known to influence construction activity. Population density and average household income represent key socioeconomic indicators, while budget allocations for Growth-Related, State of Good Repair, and Service Improvement projects drive infrastructure investments.

During causal analysis, population density exhibited unexpected exclusion restriction characteristics, showing a negative relationship with building permits. To enhance interpretability, population density was excluded from the model, leaving average household income and budget allocations as the primary predictors. The refitted model focuses on understanding how economic and infrastructural factors shape construction activity.

# Results {#sec-results}

In this section, we visualize the relationships between population density, average household income, budget allocations with total number of building permits across Toronto's 25 wards. Additionally, we present our model results, highlighting how these variables influence the number of building permits issued in each ward.

## Distribution of Average Household Income and Population Density {#sec-distribution}

The City of Toronto is divided into 25 wards, and the 2021 census data highlights significant disparities in population and average household income among them. As shown in [@fig-toronto-map-population], Spadina-Fort York (Ward 10) and Etobicoke-Lakeshore (Ward 3) have the highest population densities, with 135,400 and 139,920 residents, respectively. In contrast, Don Valley West (Ward 15) has a lower population density of 101,025 but boasts the highest average household income at \$224,800.

Population distribution presents a concentration in the western part of the city and downtown areas. Notably, wards like Etobicoke Centre (Ward 2) and Etobicoke-Lakeshore (Ward 3), situated farther from downtown, have lower costs of living and correspondingly lower average household incomes, such as Etobicoke North (Ward 2) with \$95,200.

An interesting pattern emerges around Don Valley West (Ward 15), which has the highest average household income. Wards such as University-Rosedale (Ward 11), Toronto-St. Paul's (Ward 12), and Eglinton-Lawrence (Ward 8) are clustered together, indicating a geographical correlation among high-income areas. Conversely, the wards with the highest population densities do not overlap with those that have the highest average household incomes, highlighting distinct socioeconomic patterns within the city.

## Relationship between Confounders, Mediators and Budget Allocations {#sec-rel-budget}

We now examine the results of independent comparisons between the outcome variable (total number of building permits in Toronto's 25 wards) and the predictor variables: population density, average household income, and budget allocations for Growth-Related, State of Good Repair, and Service Improvement and Enhancement projects. As illustrated in [@fig-building-permits-relationship], most predictor variables demonstrate a strong positive correlation with the number of building permits, suggesting that higher population density, average household income, and budget allocations are generally associated with increased construction activity and urban development.

However, the role of population density presents a different narrative. Although [@sec-rel-variables] revealed a positive correlation between population density and building permits, a counterintuitive negative relationship emerged within the model, leading to its exclusion. This anomaly is largely explained by the influence of Ward 10 (Spadina-Fort York) and Ward 3 (Etobicoke-Lakeshore), which consistently exhibited high population densities, substantial budget allocations, and elevated building permit counts. 

When these two wards are excluded, a clearer pattern emerges, Wards 8 (Eglinton-Lawrence), 11 (University-Rosedale), 2 (Etobicoke Centre), and 15 (Don Valley West) dominate in terms of building permits, average household income, and budget allocations. These findings suggest that, while population density plays a role in shaping construction activity, its effects are overshadowed by the more direct influence of income levels and budget distribution, warranting its exclusion to reduce model distortion and improve interpretability.


```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-building-permits-relationship
#| fig.cap: "Relationship between Variables and Total Building Permits"
#| fig-subcap: ["Population", "Average Household Income", "Total Budget in Three Categories"]
#| fig-pos: "h"
#| layout-ncol: 2

# Create multiple scatter plots
plot1 <- ggplot(results_data, aes(x = population, y = total_building_permits, label = ward_id)) +
    geom_point(color = "blue") +
    geom_text_repel(hjust = -0.5, vjust = 0.5) +
    geom_smooth(method = "lm", color = "red", se = TRUE) +
    labs(x = "Population", y = "Total Building Permits")

plot2 <- ggplot(results_data, aes(x = average_household_income, y = total_building_permits, label = ward_id)) +
    geom_point(color = "blue") +
    geom_text_repel(hjust = -0.5, vjust = 0.5) +
    geom_smooth(method = "lm", color = "red", se = TRUE) +
    labs(x = "Average Household Income", y = "Total Building Permits")

plot3 <-
  ggplot(results_data, aes(x = total_budget, y = total_building_permits, label = ward_id)) +
    geom_point(color = "blue") +
    geom_text_repel(hjust = -0.5, vjust = 0.5) +
    geom_smooth(method = "lm", color = "red", se = TRUE) +
    labs(
      x = "Total Budget in Three Categories (Sum)",
      y = "Total Building Permits",
    )

plot1
plot2
plot3
```

## Model Results {#sec-model-results}

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-model-results
#| tbl-cap: "Model Results"
#| tbl-pos: "h"

# model summary
modelsummary(
  list(
    "Model" = model
  )
)
```

The Bayesian regression model results, summarized in [@tbl-model-results], provide understanding into the relationship between budget allocations, average household income, and building permits across Toronto’s wards. The model explains 59.1% of the variance in building permits, with significant positive coefficients for total budget (0.525) and average household income (0.365). This indicates that wards with higher budgets and income levels are more likely to experience construction activity.

Ward-specific random effects shows the influence of unobserved, ward-specific factors on permit issuance, while the adjusted ($R^2$) of 0.359 suggests moderate explanatory power after accounting for predictor complexity. With an RMSE of 0.57, the model achieves reasonable predictive accuracy.

While the assumption of linearity and additive effects simplifies the analysis, it may overlook variable interactions. The exclusion of population density, due to its counterintuitive negative relationship, reflects a trade-off between clarity and capturing complex dynamics. Despite these limitations, the findings highlight the critical role of economic resources—both income and budget—in driving urban development, emphasizing the need for equitable resource allocation in shaping infrastructure growth.

# Discussion {#sec-discussion}

In this section, we discuss the implications of our findings on urban development, economic well-being, and resource distribution in Toronto. We explore the role of budget allocations, average household income, and population density in shaping construction activity and infrastructure development across the city's 25 wards. Our analysis highlights the complex interplay between economic factors, demographic trends, and policy decisions, shedding light on the challenges and opportunities for equitable urban growth in Toronto.

## Building Permits and Economic Well-Being

Construction plays a pivotal role in the economy, driving both job creation and economic growth [@citeConstruction]. One of the key indicators of construction activity is the number of building permits issued within a ward. Increased construction correlates with job opportunities, infrastructure development, and overall community well-being. For example, investment in residential and non-residential construction rose from 24.9% in 2021 to 29.5% in 2023, with projections indicating a further 1.5% increase in construction employment by 2024 [@citeConstructionEmployment].

As highlighted in [@sec-model-results] Bayesian regression model shows that average household income has a significant positive effect on the issuance of building permits. This suggests that construction activity tends to concentrate in higher-income areas, where demand for development is stronger. This finding underscores the role of income levels in driving construction and infrastructure development, emphasizing the need for targeted investments in lower-income areas to stimulate economic growth and enhance community well-being.

For potential property investors in Toronto, these understanding offer a strategic advantage. Investing in areas with higher average household incomes may be more profitable due to the increased likelihood of ongoing construction activity, which can drive property value appreciation and strengthen the local economy. While high-population-density areas are typically attractive for investment, our findings suggest that this may not always be the case.

## Investing in Higher-Income Areas vs. High-Density Areas

The interplay between population density, average household income, and infrastructure investment in Toronto highlights a complex challenge in achieving equitable resource distribution. As seen in [@sec-rel-variables] high-income areas often attract more construction activity, driven by market forces and residents’ ability to finance new developments. Whereas, only specific high-density areas like Ward 10 (Spadina-Fort York) and Ward 13 (Toronto Centre) receive substantial investment despite high population density, justifying the reason for population density exclusion in the model. Meanwhile, as discussed in [@sec-distribution], low-density and lower-income wards risk being underserved.

These findings underscore the importance of equity-based budgeting, where low-density and lower-income areas should receive targeted investments to meet infrastructure needs. This approach aligns with urban development strategies that prioritize inclusivity. Notable examples include transit expansions such as the Ontario Line and Crosstown extensions [@citeTransitExpansion].The Ontario Line is expected to connect high-density areas like Ward 10 (Spadina-Fort York) and Ward 13 (Toronto Centre) to the broader transit network, which could stimulate further development and infrastructure investment in these wards, fostering a more balanced urban landscape and distributing population growth more evenly across the city.

For future policy, Toronto could enhance equity by expanding initiatives such as eliminating parking minimums. This policy reduces construction costs, making development in high-density areas more feasible. Such measures would promote a more balanced and sustainable urban environment, ensuring that population growth is accompanied by appropriate infrastructure investments.


## Budget Allocations and Construction Activity

Our analysis shows a complex relationship between budget allocations and construction activity in Toronto's 25 wards. As discussed in [@sec-budget-trends] there has been a steady increases in budget allocations for key initiatives such as Growth-Related, State of Good Repair, and Service Improvement and Enhancement projects from 2021 to 2024, the distribution of building permits shows only a good correlation with these allocations [@sec-model-results]. This finding suggests that budget allocations alone may not be the sole driver of construction activity, as other factors like market conditions, regulatory frameworks, and community priorities also play a significant role.

This discrepancy reflects broader challenges highlighted in Toronto's recent 10-year economic plan [@citeEconomicPlan], which aims to tackle pressing issues such as housing, inequality, and congestion. The plan emphasizes the need for strategic investments that go beyond conventional infrastructure spending. For example, the city has committed $35 million to cultural initiatives over the next decade, signaling a shift toward holistic urban development that prioritizes cultural and social well-being alongside physical infrastructure improvements. This shift may explain the weak correlation between budget allocations and building permits, as construction activity is influenced by a broader range of factors beyond traditional infrastructure spending.

Moreover, the Ford government’s fall economic statement indicates a shrinking provincial deficit, suggesting a cautious fiscal environment that may limit how budgetary increases translate into tangible construction outputs [@citeFallEconomicStatement]. These dynamics suggest that other factors, such as regulatory frameworks, market conditions, or demographic shifts, may be playing a more significant role than budgetary allocations alone in shaping Toronto’s construction landscape.


# Weaknesses and Future Work {#sec-limitations}

This analysis has several limitations that warrant consideration. As discussed in [@sec-data], the datasets used, sourced from the City of Toronto, provide insights into the city’s demographics, budget allocations, and building permits. However, they may not fully capture the complexity of Toronto's infrastructure development and resource distribution. 

For instance, in [@sec-building-permits], all building *Work Types* were included in the analysis, which may not accurately represent construction activity in each ward. Many recorded work types, such as renovations, demolitions, and minor repairs (e.g., HVAC installations or pipework), are not indicative of significant construction projects. Future research could refine the analysis by focusing on specific types of construction projects to better reflect actual development activity.

Additionally, as noted in [@sec-appendix], there are 45 unique statuses assigned to building permits, but this study analyzed only a subset of these statuses. This narrow focus may overlook important stages in the construction process. Expanding the analysis to include more permit statuses could offer a more comprehensive understanding of construction activity.

Another limitation is the lack of detailed data on specific projects within each budget category, which restricts insights into why certain wards receive more or less funding. Access to granular project-level data could provide deeper context and explanations for funding disparities.

Moreover, the analysis assumes linear relationships between population density, average household income, and budget allocations, which may oversimplify the underlying dynamics. Developing more complex models that account for non-linear relationships and interactions could yield a richer understanding of factors influencing construction activity. 

Finally, the analysis assumes independence among wards, an assumption that may not hold true given shared infrastructure and resource dependencies. Future studies could incorporate spatial or network-based models to address interdependencies between wards.

## Future Work

This analysis provides a foundation for future research on infrastructure development, resource distribution, and economic well-being in Toronto. Future studies could explore additional factors that influence budget allocations and building permit issuance, such as infrastructure needs, community priorities, and political considerations. By incorporating these variables into the analysis, researchers can gain a more in-depth understanding of how municipal budgets are allocated and how construction activity is distributed across the city's wards.

In terms of policy recommendations, the city should adopt equity-based budgeting to ensure underfunded, high-density wards receive adequate resources, particularly for essential services like health and safety. Toronto could also implement a baseline funding model for high-need areas and enhance public engagement in the budget process, ensuring that residents of under served wards have a voice in resource allocation. These actions would promote more transparent, balanced, and equitable urban governance.

\newpage

\appendix

# Appendix {#sec-appendix}

The relationship between income levels, population density, and construction outcomes reflects complex interactions shaped by diverse socioeconomic and regional factors. Existing datasets, such as the NHS survey, provide a starting point but do not fully capture these dynamics. To address this, a well-structured survey methodology is outlined to ensure greater representativeness and reliability.

## Survey Design for Construction Analysis

The survey seeks to understand how income levels influence construction investments, access to resources, regulatory challenges, and housing quality across geographic regions. This allows for a detailed assessment of regional and economic differences in construction outcomes.  

## Sampling Methodology
The proposed design prioritizes representativeness through the following methods:  

- Stratified Sampling: Groups are categorized by income levels and ward classifications (high-income, populated, and etc.) to ensure diverse representation.  
- Random Sampling Within Strata: This approach minimizes selection bias, ensuring participants across income groups are adequately represented.  

## Ensuring Robust Data Collection  

- Sampling Bias: Using stratified random sampling ensures inclusion of marginalized groups, particularly those in remote wards or lower income brackets, often underrepresented in traditional data sources.  
- Non-Response Bias: Offering multiple delivery modes—such as online forms, mailed surveys, and in-person interviews—accommodates varying levels of accessibility. To encourage participation, incentives like gift cards or small monetary rewards can be provided.  
- Question Design: Survey questions must be neutrally worded to avoid influencing responses. Pilot testing and iterative feedback from participants will refine question clarity and relevance.  
- Human-Centric Approach: Present the survey as an opportunity for participants to share their experiences and shape future construction planning.  
- Visual Enhancements: Include infographics or interactive components to make participation more engaging and relatable.  
- Validation Through Multiple Sources: To strengthen findings, the survey data should be cross-referenced with secondary sources, such as municipal reports, and supplemented with qualitative interviews where necessary.  

This approach ensures a structured methodology that balances statistical rigor with participant-focused design, allowing for a meaningful exploration of how socioeconomic factors influence construction outcomes across regions.

## Building Permits Status

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-status-building-permits-data
#| tbl-cap: "Building permit statuses"
#| tbl-pos: H

# Display the unique statuses in the building permits data using kable
building_permits |>
  distinct(STATUS) |>
  kable() |>
  kable_styling(full_width = F, position = "center")

```

## Model Details {#sec-model-details}

## Posterior Predictive Check

We employ a posterior predictive check in @fig-ppcheck to evaluate how well the model fits the data. By examining the distribution of the observed data against the simulated data, we can assess the model's predictive accuracy and identify any discrepancies. [@citeAlexander].

```{r}
#| label: fig-ppcheck
#| fig-cap: Examining how the model fits the data through a posterior predictive check
#| echo: false
#| fig-pos: H
#| message: false
#| warning: false

pp_check(model) + 
  theme_classic() +
  theme(legend.position = "bottom")
```

## Comparison of the Posterior and Prior

We compare the posterior and prior in @fig-postprior to examine how the estimates change once data is taken into account [@citeAlexander].

```{r}
#| label: fig-postprior
#| fig-cap: Examining how the model is affected by the data through a posterior-prior comparison
#| echo: false
#| warning: false
#| fig-pos: H
#| message: false
#| fig-width: 10
#| fig-height: 10

model_data <- analysis_budget_data |>
  group_by(ward_id, category, ward) |>
  summarise(avg_total_10_year = mean(total_10_year)) |>
  pivot_wider(names_from = category, values_from = avg_total_10_year) |>
  # sum the total 10 year budget for each ward
  group_by(ward_id, ward) |>
  summarise(total_budget = sum(`Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`))

# combine the model data with the ward profile data
model_data <- model_data |>
  left_join(analysis_ward_data, by = "ward_id") |>
  select(ward_id, ward.x, total_budget, average_household_income, population) |>
  rename(ward = ward.x) |>
  # left join with the building permits data
  left_join(analysis_building_permits, by = "ward_id") |>
  select(ward_id, ward, total_budget, average_household_income, population, total_building_permits)


posterior_vs_prior(model) + 
  theme_classic() + 
  theme(legend.position = "bottom") +
  coord_flip()
```

## Markov Chain Monte Carlo Convergence


Since `rstanarm` uses the MCMC sampling algorithm [@citeAlexander], we assess whether the algorithm encountered any issues by reviewing the Rhat and trace plots, as shown in [@fig-mcmc]. The Rhat plot shows no concerns, with values consistently close to 1, indicating that the algorithm converged properly [@citeAlexander]. Additionally, the trace plot appears normal, as the lines remain horizontal and fluctuate as expected, further supporting the adequacy of the sampling process [@citeAlexander].

```{r}
#| label: fig-mcmc
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Rhat plot", "Trace plot"]
#| echo: false
#| message: false
#| warning: false
#| fig-pos: H
#| layout-ncol: 1
#| fig-width: 10
#| fig-height: 5

plot(model, "rhat") + 
  theme_classic()

plot(model, "trace") 

```

## Credibility Intervals

Lastly, we employ a 95% credibility interval as showcased in @fig-ci, to gain a better understanding of the probability distribution of our coefficients.

```{r}
#| label: fig-ci
#| fig-cap: "Credible intervals"
#| echo: false
#| message: false
#| warning: false
#| fig-pos: H

plot(model) + 
  theme_classic()

```

\newpage
# References
